The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-23 conf_article/neon.tex
Added expressions for f2 and f* duals

        where $\alpha \in [0, 1]$ defines a specific norm from the $F*$-family.

@@

        where $\alpha \in [0, 1]$ defines a specific norm from the F2-family.

@@

Despite such similarities, no theory has been yet proposed that would reduce the matrix case to the optimization of a function of a singular numbers vector. If such a theory is developed, analysis of matrix algorithms like Muon will be greatly simplified.

@@

\subsection{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
From \cite[p. 2]{yu2012arithmetic} it follows that the norm ball in $\normfstar{\cdot}^\dagger$ is the Minkowski sum of the norm ball in $\alpha\normn{\cdot}$ and $(1-\alpha)\normf{\cdot}$. The norm could be reconstructed from this geometric observation.

Similarly, the norm ball in $\normftwo{\cdot}^\dagger$ is the Minkowski sum of the norm ball in $\alpha\norms{\cdot}$ and $(1-\alpha)\normf{\cdot}$.

\subsection{Visualization of different matrix norms}
\subsubsection{Duals to F* and F2 norms}
In \ref{fig:fduals} we plot these norms. On x-axis and y-axis, there are singular values $\sigma_1$, $\sigma_2$ respectively of a matrix from $\Rmn$ with $\min\{m,n\}=2$.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fstardualball.pdf}
        \caption{The dual to F* norm for different $\alpha$}
        \label{fig:fstardual}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{ftwodualball.pdf}
        \caption{The dual to F2 norm for different $\alpha$}
        \label{fig:ftwodual}
    \end{subfigure}
    \caption{The duals to F* and F2 norms for different $\alpha$}
    \label{fig:fduals}
\end{figure}

@@

To showcase the complex structure of the Ky Fan $k$-norm and its dual, we suggest the illustrations \ref{fig:kyfan_combined} with the ball in the Ky Fan $2$-norm in \ref{fig:kyfan} and its dual in \ref{fig:kyfandual}. On x-, y-, and z-axes, there are singular values $\sigma_1$, $\sigma_2$, and $\sigma_3$ respectively of a matrix from $\Rmn$ with $\min\{m,n\}=3$. In this particular case, we do not sort the singular values. In the proposed representation, we actually plot balls Top-$2$ norm $\max\{\abs{x}+\abs{y}, \abs{x}+\abs{z}, \abs{y}+\abs{z}\}$ and its dual norm $\max\{\max(\abs(x),\abs{y},\abs{z}), \frac{1}{2} (\abs{x}+\abs{y}+\abs{z})\}$. The resulting balls are much more complex than balls in $l_\infty$, $l_1$ and $l_2$ norms.

>>>>

        where $\alpha \in [0, 1]$ defines a specific norm from the $F*$-family. $\normfstar{\cdot}^\dagger$ can be expressed as in \ref{eq:normfstardual}. However, its norm ball is a simple Minkowski sum of $\alpha \normn{\cdot}$ and $(1-\alpha)\normf{\cdot}$ balls \ref{fig:fstardual}.

@@

        where $\alpha \in [0, 1]$ defines a specific norm from the F2-family.$\normftwo{\cdot}^\dagger$ can be expressed as in \ref{eq:normftwodual}. However, its norm ball is a simple Minkowski sum of $\alpha \norms{\cdot}$ and $(1-\alpha)\normf{\cdot}$ balls \ref{fig:ftwodual}.

@@

Despite such similarities, no theory has been yet proposed that would reduce the matrix case to the optimization of a function of a singular values vector. If such a theory is developed, analysis of matrix algorithms like Muon will be greatly simplified.

@@

\subsection{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
First, we need a well-known fact mentioned, for example, in \cite[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof of the fact.

\begin{lemma}\label{lem:dual_to_conv_comb}
    Let $\norm{\cdot}_{(1)}$ and $\norm{\cdot}_{(2)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha,\beta \geq 0$. Define 
    $$
    \norm{x} := \alpha \norm{x}_{(1)} + \beta \norm{x}_{(2)}.
    $$
    Then the dual unit ball of $\norm{\cdot}$ satisfies
    $$
    B_{\norm{\cdot}^\dagger} 
    = \alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger},
    $$
    where $+$ denotes the Minkowski sum and $B_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

    \begin{proof}
    Write $f(x) = \alpha \norm{x}_{(1)}$ and $g(x) = \beta \norm{x}_{(2)}$, so 
    $$
    \norm{x} = f(x)+g(x).
    $$
    Recall two standard facts:
    \begin{enumerate}
        \item For any norm $\norm{\cdot}$ and $\lambda>0$, 
        $$
        (\lambda \norm{\cdot})^*(y) = 
        \sup_{x}\bigl( \langle y,x \rangle - \lambda \norm{x}\bigr)
        = \delta_{\lambda B_{\norm{\cdot}^\dagger}}(y),
        $$
        i.e., the indicator function of the scaled dual ball.
        \item The Fenchel conjugate of a sum satisfies
        $$
        (f+g)^*(y) = \inf_{u+v=y} \bigl(f^*(u) + g^*(v)\bigr).
        $$
    \end{enumerate}
    Applying these to $f$ and $g$, we have 
    $$
    f^*(u) = \delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}}(u), 
    \quad 
    g^*(v) = \delta_{\beta B_{\norm{\cdot}_{(2)}^\dagger}}(v).
    $$
    Thus
    $$
    \norm{\cdot}^*(y) 
    = (f+g)^*(y)
    = \inf_{u+v=y}
    \bigl(
    \delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}}(u)
    +
    \delta_{\beta B_{\norm{\cdot}_{(2)}^\dagger}}(v)
    \bigr)
    = 
    \delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}+\beta B_{\norm{\cdot}_{(2)}^\dagger}}(y).
    $$
    But by definition, the conjugate of a norm is exactly the indicator of its dual unit ball:
    $$
    \norm{\cdot}^*(y) = \delta_{B_{\norm{\cdot}^\dagger}}(y).
    $$
    Therefore,
    $$
    B_{\norm{\cdot}^\dagger} = \alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger}.
    $$
    \end{proof}

Consequently,
\begin{equation}\label{eq:dual_to_conv_comb}
\norm{y}^\dagger
= \inf\bigl\{ t \ge 0 : y \in t\bigl(\alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger}\bigr) \bigr\}
= \inf_{z,t}\bigl\{t \ge 0: z \in t \alpha B_{\norm{\cdot}_{(1)}^\dagger}, y - z \in t \beta B_{\norm{\cdot}_{(1)}^\dagger}\bigr\}
\end{equation}

Thus, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get 
\begin{equation}\label{eq:normfstardual}
\normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

If $\alpha = 1$, then $\mZ = \mY$, and we get $\normfstar{\mY}^\dagger = \norms{\mY}$. If $\alpha = 0$, then $\mZ = 0$, and we get $\normfstar{\mY}^\dagger = \normf{\mY}$.

Similarly, we find $\normftwo{\cdot}^\dagger$, which is related to F-Neon update:
\begin{equation}\label{eq:normftwodual}
\normftwo{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \normn{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

If $\alpha = 1$, then $\mZ = \mY$, and we get $\normftwo{\mY}^\dagger = \normn{\mY}$. If $\alpha = 0$, then $\mZ = 0$, and we get $\normftwo{\mY}^\dagger = \normf{\mY}$.

\subsection{Visualization of different matrix norms}
\subsubsection{Duals to F* and F2 norms}

 It follows from \ref{lem:dual_to_conv_comb} that the norm ball in $\normfstar{\cdot}^\dagger$ is the Minkowski sum of the norm ball in $\alpha\normn{\cdot}$ and $(1-\alpha)\normf{\cdot}$ and the norm ball in $\normftwo{\cdot}^\dagger$ is the Minkowski sum of the norm ball in $\alpha\norms{\cdot}$ and $(1-\alpha)\normf{\cdot}$.

In \ref{fig:fduals} we plot these norms. On x-axis and y-axis, there are singular values $\sigma_1$, $\sigma_2$ respectively of a matrix from $\Rmn$ with $\min\{m,n\}=2$.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fstardualball.pdf}
        \caption{lmo balls for F-Muon for different $\alpha$}
        \label{fig:fstardual}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{ftwodualball.pdf}
        \caption{lmo ball for F-Neon for different $\alpha$}
        \label{fig:ftwodual}
    \end{subfigure}
    \caption{Balls in the duals to F* and F2 norms for different $\alpha$}
    \label{fig:fduals}
\end{figure}

@@

To showcase the complex structure of the Ky Fan $k$-norm and its dual, we suggest the illustrations \ref{fig:kyfan_combined} with the ball in the Ky Fan $2$-norm in \ref{fig:kyfan} and its dual in \ref{fig:kyfandual}. On x-, y-, and z-axes, there are singular values $\sigma_1$, $\sigma_2$, and $\sigma_3$ respectively of a matrix from $\Rmn$ with $\min\{m,n\}=3$. In this particular case, we do not sort the singular values. In the proposed representation, we actually plot balls in the Top-$2$ norm $\max\{\abs{x}+\abs{y}, \abs{x}+\abs{z}, \abs{y}+\abs{z}\}$ and its dual norm $\max\{\max(\abs(x),\abs{y},\abs{z}), \frac{1}{2} (\abs{x}+\abs{y}+\abs{z})\}$. The resulting balls are much more complex than balls in $l_\infty$, $l_1$ and $l_2$ norms.
